### 2024-09-16 manuscripts/07-economic-letters-v5/Manuscript.tex
submit to economic letters

\begin{abstract}
\singlespacing % Set single spacing for the abstract
Do individuals with Hispanic names face labor market discrimination? This study analyzes the impact of Hispanic-sounding surnames on wages among inter-ethnic children with one White and one Hispanic parent. I find that individuals with Hispanic surnames receive less education, more likely to be unemployed, and often earn less, with a notable wage gap favoring those with White surnames. People born to Hispanic fathers and White mothers receive 0.3 years of education less than those born to White fathers and Hispanic mothers. Males born to Hispanic fathers and White mothers are 1 percentage point more likely to be unemployed, and they also earn 5 percentage points less than those born to White fathers and Hispanic mothers. The earnings gap is largely due to educational differences.
\end{abstract}

@@

A substantial earnings gap exists between native-born White Hispanic men and White men, with Hispanic men earning 21\% less. While a significant portion of this gap is attributed to educational differences, discrimination may also play a role, potentially leading to reduced job opportunities, lower wages, and hindered assimilation \citep{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}. This paper examines the impact of Hispanic last names on labor market and educational outcomes, contributing to the literature by isolating the effect of ethnicity-signaling surnames from other potentially confounding factors. By focusing on children of inter-ethnic marriages, this study provides a novel approach to understanding the mechanisms behind Hispanic-White earnings disparities and the potential role of name-based discrimination in the labor market.

@@

I find that individuals with Hispanic-sounding last names receive 0.2 years less education and are more likely to be unemployed. The gap in education is equivalent to a 1.9\% reduction, and the unemployment gap is equal to a 14\% increase. The earnings gap between individuals with Hispanic-sounding last names and those with White last names is 5 percentage points. However, when controlling for education, the earnings gap reduces to 1 percentage point, suggesting that the earnings gap is largely due to differences in education.

@@

The CPS sample is restricted to White, U.S.-born citizens aged 25-40, born between 1950 and 2000. Parents are classified as Hispanic if born in a Spanish-speaking country or Puerto Rico, and White if they are US-born. Four parental types are identified: (1) White father and White mother (WW), (2) White father and Hispanic mother (WH), (3) Hispanic father and White mother (HW), and (4) Hispanic father and Hispanic mother (HH). Table \ref{tab:combined-mat} shows the distribution of these types. WW children comprise 96\% of the sample, HH 3\%, and inter-ethnic children (WH and HW) 1.35\%. Summary statistics for each group are presented in Table \ref{tab:combined-sum}. Using the 1950-2000 censuses, I constructed a dataset of synthetic parents, including married White men and women born between 1920-1975.\footnote{To proxy for parental characteristics, I generate `synthetic' parents, assuming mothers were between the ages of 20 and 35 at the time of birth. For example, a `synthetic mother' of a 35-year-old in 2019, with a father born in the United States and a mother born in Colombia, would have the characteristics of an average married woman born in Colombia between 1949 and 1964, based on both her and her spouse's birthplaces. This family would appear in the 1990 Census with a 6-year-old child, with the mother born in Colombia and the father being U.S.-born.} Hispanic individuals are defined as White and born in a Spanish-speaking country, while Whites are native-born. Table \ref{tab:combined-mat} shows the distribution of couple types: WW (96\%), HH (2\%), WH (<1\%), and HW (1\%). Summary statistics for the parents are presented in Table \ref{tab:combined-sum}.

@@

I estimate the effect of having a Hispanic last name on educational outcomes and earnings. The difference in means between Hispanics and non-Hispanic Whites could result from either discrimination or differences in innate abilities, skills, and parental investments. While controlling for observable skill measures, I compare children of inter-ethnic marriages, HW and WH. WH and HW children are more similar in characteristics but provide employers, and the labor market, with different signals.\footnote{WH and HW children are both half White, half Hispanic.} WH children will have a non-Hispanic last name, while HW children will have a Hispanic last name. This is a method developed by \citet{rubinstein2014pride}.

@@

In this section I go over my empirical approach. I restrict the sample to WH and HW groups. Let $Y_{ist}$ be the outcome of interest for person $i$ in state $s$ at time $t$. $HW_{ist}$ is an indicator variable for the type of parents person $i$ has. $X_{ist}$ is a vector of controls that includes age and numbers of hours worked, $\gamma_{st}$ are state-year fixed effects (FE), and $\phi_{ist}$ represents the error term. The equation for this strategy is written as follows:
\begin{equation} \label{eq:1a}
Y_{ist} = \beta_{1} HW_{ist} + X_{ist} \pi + \gamma_{st} + \phi_{ist}
\end{equation}

$\beta_{1}$ is the coefficient of interest in this specification. $\beta_{1}$ represents the gaps in outcomes between children of inter-ethnic marriages who have a Spanish-sounding last name versus a native-sounding last name.

@@

I show in Table \ref{tab:lastnamereg} a 1 percentage point employment gap between HW and WH workers, and a 5 percentage point crude earnings gap. However, when controlling for education, the earnings gap reduces to a statistically insignificant 1 percentage point. I also find that HW children are 1 percentage point more likely to be unemployed than WH children. The gap in unemployment is equivalent to a 14\% increase and is statistically significant even after controlling for education. These results suggest that having a Hispanic last name negatively impacts labor market outcomes, particularly in terms of employment.

@@

These results align with existing literature on disparities in educational access \citep{bergman2018education,gaddis2024racial}. While the earnings gap diminishes when controlling for education, this doesn't necessarily indicate an absence of discrimination, as education itself can be influenced by bias. Moreover, educational differences cannot explain the gaps in unemployment, suggesting that discrimination may play a role in labor market outcomes of individuals with Hispanic-sounding last name. Further research is needed to comprehensively understand the earnings gaps between Hispanics and Whites, particularly considering that mothers of HW children have higher education and earnings, which theoretically should lead to better outcomes for their children.

@@

\bibliographystyle{cas-model2-names}
\bibliography{bibliography}
\newpage
\clearpage
%table1
\input{tables/tab16-tab1-and-tab2}

@@

%table3
\input{tables/tab17-all-sums}

@@

\input{tables/tab13-regression}

@@

\input{tables/tab15-emp-earn-regression.tex}

>>>>

\begin{abstract}
\singlespacing % Set single spacing for the abstract
Do individuals with Hispanic names face labor market discrimination? This study analyzes the impact of Hispanic-sounding surnames on wages among inter-ethnic children with one White and one Hispanic parent. I find that individuals with Hispanic surnames receive less education, more likely to be unemployed, and often earn less, with a notable wage gap favoring those with native-sounding surnames. People born to Hispanic fathers and White mothers receive 0.3 years of education less than those born to White fathers and Hispanic mothers. Males born to Hispanic fathers and White mothers are 1 percentage point more likely to be unemployed, and they also earn 5 percentage points less than those born to White fathers and Hispanic mothers. The earnings gap is largely due to educational differences.
\end{abstract}

@@

A substantial earnings gap exists between native-born White Hispanic men and White men, with Hispanic men earning 21\% less. While a significant portion of this gap is attributed to educational differences, discrimination may also play a role, potentially leading to reduced job opportunities, lower wages, and hindered assimilation \citep{duncan2006hispanics, duncan2018identifying, duncan2018socioeconomic}. This paper examines the impact of Hispanic last names on labor market and educational outcomes, contributing to the literature by isolating the effect of ethnic-sounding surnames from other potentially confounding factors. By focusing on children of inter-ethnic marriages, this study provides a novel approach to understanding the mechanisms behind Hispanic-White earnings disparities and the potential role of name-based discrimination in the labor market.

@@

I find that individuals with Hispanic-sounding last names receive 0.2 years less education and are more likely to be unemployed. The gap in education is equivalent to a 1.5\% reduction, and the unemployment gap is equal to a 14\% increase. The earnings gap between individuals with Hispanic-sounding last names and those with White last names is 5 percentage points. However, when controlling for education, the earnings gap decreases to 1 percentage point, suggesting that the earnings gap is largely due to differences in education.

@@

The CPS sample is restricted to White, U.S.-born citizens aged 25-40, born between 1950 and 2000. Parents are classified as Hispanic if born in a Spanish-speaking country or Puerto Rico, and as White if they are US-born non-Hispanics. Four parental types are identified: (1) White father and White mother (WW), (2) White father and Hispanic mother (WH), (3) Hispanic father and White mother (HW), and (4) Hispanic father and Hispanic mother (HH). Table \ref{tab:combined-mat} shows the distribution of these types. WW children comprise 96\% of the sample, HH 3\%, and inter-ethnic children (WH and HW) 1.35\%. Summary statistics for each group are presented in Table \ref{tab:combined-sum}. Using the 1950-2000 censuses, I constructed a dataset of synthetic parents, including married White men and women born between 1920-1975.\footnote{To proxy for parental characteristics, I generate `synthetic' parents, assuming mothers were between the ages of 20 and 35 at the time of birth. For example, a `synthetic mother' of a 35-year-old in 2019, with a father born in the United States and a mother born in Colombia, would have the characteristics of an average married woman born in Colombia between 1949 and 1964, based on both her and her spouse's birthplaces. This family would appear in the 1990 Census with a 6-year-old child, with the mother born in Colombia and the father being U.S.-born.} Hispanic individuals are defined as White and born in a Spanish-speaking country, while Whites are native-born. Table \ref{tab:combined-mat} shows the distribution of couple types: WW (96\%), HH (2\%), WH (<1\%), and HW (1\%). Summary statistics for the parents are presented in Table \ref{tab:combined-sum}.

@@

I estimate the effect of having a Hispanic last name on educational outcomes and earnings. The difference in means between Hispanics and non-Hispanic Whites could result from either discrimination or differences in innate abilities, skills, and parental investments. While controlling for observable skill measures, I compare children of inter-ethnic marriages, HW and WH. WH and HW children are more similar in characteristics but provide employers, and the labor market, with different signals.\footnote{WH and HW children are both half White, half Hispanic.} WH children will have a non-Hispanic last name, while HW children will have a Hispanic last name.

@@

In this section, I discuss my empirical approach. I restrict the sample to WH and HW groups. Let $Y_{ist}$ be the outcome of interest for person $i$ in state $s$ at time $t$. $HW_{ist}$ is an indicator variable for the type of parents person $i$ has. $X_{ist}$ is a vector of controls that includes age and numbers of hours worked, $\gamma_{st}$ are state-year fixed effects (FE), and $\phi_{ist}$ represents the error term. The equation for this strategy is written as follows:
\begin{equation} \label{eq:1a}
Y_{ist} = \beta_{1} HW_{ist} + X_{ist} \pi + \gamma_{st} + \phi_{ist}
\end{equation}

$\beta_{1}$ is the coefficient of interest in this specification. $\beta_{1}$ represents the gaps in outcomes between children of inter-ethnic marriages who have an ethnic-sounding last name versus a native-sounding last name.

@@

I show in Table \ref{tab:lastnamereg} a 1 percentage point unemployment gap between HW and WH workers, and a 5 percentage point crude earnings gap. However, when controlling for education, the earnings gap reduces to a statistically insignificant 1 percentage point. I also find that HW children are 1 percentage point more likely to be unemployed than WH children. The gap in unemployment is equivalent to a 14\% increase and is statistically significant even after controlling for education. These results suggest that having a Hispanic last name negatively impacts labor market outcomes, particularly in terms of employment.

@@

These results align with existing literature on disparities in educational access \citep{bergman2018education,gaddis2024racial}. While the earnings gap diminishes when controlling for education, this doesn't necessarily indicate an absence of discrimination, as education itself can be influenced by bias. Moreover, educational differences cannot explain the gaps in unemployment, suggesting that discrimination may play a role in labor market outcomes of individuals with Hispanic-sounding last names. Further research is needed to comprehensively understand the earnings gaps between Hispanics and Whites, particularly considering that mothers of HW children have higher education and earnings, which theoretically should lead to better outcomes for their children.

@@

\bibliographystyle{cas-model2-names}
\bibliography{bibliography}
\newpage
\clearpage
%table1
\input{tab16-tab1-and-tab2}

@@

%table3
\input{tab17-all-sums}

@@

\input{tab13-regression}

@@

\input{tab15-emp-earn-regression.tex}
